The Three-HITs Theorem

### 2017-03-27 Paper/Paper.tex
Replace `\>` with `:`.
I find `\>` much harder to type, it's not clear what the advantage over `:` is, and also
`\>` does not work in indices, e.g., `\prod_{x \> A}`, at least for me.

\newcommand{\fpath}[0]{\> \Box \>}
\newcommand{\conc}[0]{\bullet}
\newcommand{\comp}[0]{\circ \,}

@@

For example, if $\nat$ is the type of natural numbers, the expression
$(X + \nat) \times X$ is a polynomial expression in~$X$. We write $P[X]$ to indicate a
polynomial in variable $X$. Given a type $A$, we may substutute $A$ for $X$ to obtain a
type $P[A]$, as follows:
%
\begin{align*}
  X[A] &\defeq A, &
  (P_1 \times P_2)[A] &\defeq P_1[A] \times P_2[A], \\
  C[A] &\defeq C, &
  (P_1 + P_2)[A] &\defeq P_1[A] + P_2[A].
\end{align*}
%
In a similar fashion a map $f \> A \to B$ is transformed to a map $P[f] : P[A] \to P[B]$:
%
\begin{align*}
  X[f] &\defeq f, &
  (P_1 \times P_2)[f] \, (x,y) &\defeq (P_1[f]\, x, P_2[f] \, y), \\
  C[f] &\defeq \idmap{C}, &
  (P_1 + P_2)[f](\inl{x}) &\defeq \inl{(P_1[f]\,x)}, \\
  & &
  (P_1 + P_2)[f](\inr{y}) &\defeq \inr{(P_2[f]\,y)}.
\end{align*}
%
Somewhat less obviously, a polynomial acts on a type family $B : A \to \Type$ to
give its lifting $\lift{P}[B] : P[A] \to \Type$,
%
\begin{align*}
  \lift{X}[B] \, u &\defeq B\,u, &
  \lift{P_1 \times P_2}[B] \, u \defeq (\lift{P_1}[B]\,u) \times (\lift{P_2}[B]\,u), \\
  \lift{C}[B] \, u &\defeq C, &
  \lift{P_1 + P_2}[B] \, (\inl{u})\defeq \lift{P_1}[B]\,u, \\
  & &
  \lift{P_1 + P_2}[B] \, (\inr{u})\defeq \lift{P_2}[B]\,u,
\end{align*}
%
while a map $f \> \dprod{x \> A} B_1\,x \to B_2\,x$ lifts to
$\lift{P}[f] \> \dprod{u : P[A]} \lift{P}[B_1]\,u \to \lift{P}[B_2]\,u$:
%
\begin{align*}
  \lift{X}[f] \, u &\defeq f \, u &
  \lift{P_1 \times P_2}[f] \, (u_1, u_2) &\defeq (\lift{P_1}[f]\,u_1, \lift{P_2}[f]\,u_2) \\
  \lift{C}[f] \, u &\defeq u &
  \lift{P_1 + P_2}[f] \, (\inl{u})\defeq \lift{P_1}[f]\,u, \\
  & &
  \lift{P_1 + P_2}[f] \, (\inr{u})\defeq \lift{P_2}[f]\,u.
\end{align*}
%
There is of course a strong analogy between the action of $P[X]$ and a polynomial functor,
but we hesitate to call $P[X]$ a functor as there is no category to speak of, and in any
case the manipulations we have just described are syntactic.

Next we define the notion of a constructor term.
\begin{definition}
\label{def:constructor-term}
Given is a function $c : A \> T \rightarrow T$ with $A$ polynomial and $T$ a type.
Then we say $t$ is a \emph{constructor term} over $c$ if we can find polynomials $F$ and $G$ such that $x : F \> T \Vdash t : G \> T$ can be derived using the following rules.
\begin{gather*}
        \AxiomC{$t : B$}
        \AxiomC{$T$ does not occur in $B$}
        \BinaryInfC{$x:F \> T \Vdash t : B$}
        \bottomAlignProof
        \DisplayProof 
         \quad
        \AxiomC{}
        \UnaryInfC{$x : F \> T \Vdash x : F \> T$}
        \bottomAlignProof
        \DisplayProof
        \quad
        \AxiomC{$x : F \> T \Vdash r : A \> T$}
        \UnaryInfC{$x : F \> T \Vdash c \> r : T$}
        \bottomAlignProof
        \DisplayProof
        \\[7pt]
        \AxiomC{$j \in \{1, 2\}$}
        \AxiomC{$x : F \> T \Vdash r : G_1 \> T \times G_2 \> T$}
        \BinaryInfC{$x : F \> T \Vdash\pi_j \> r : G_j \> T$}
        \bottomAlignProof
        \DisplayProof
        \quad
        \AxiomC{$j \in \{1, 2\}$}
        \AxiomC{$x : F \> T \Vdash r_j : G_j$}
        \BinaryInfC{$x : F \> T \Vdash (r_1, r_2) : G_1 \> T\times G_2 \> T$}
        \bottomAlignProof
        \DisplayProof
        \\[7pt]
        \AxiomC{$j \in \{1, 2\}$}
        \AxiomC{$x : F \> T \Vdash r : G_j \> T$}
        \BinaryInfC{$x : F \> T : \inn_j \> r : G_1 \> T + G_2 \> T$}
        \bottomAlignProof
        \DisplayProof
\end{gather*}
\end{definition}

Using constructor terms we give the following scheme of higher inductive types.
\begin{definition}
\label{def:hit}
A \emph{higher inductive type} is defined according to the following scheme
\lstset{language=Coq}
\begin{lstlisting}
Inductive $H$ :=
| $c$ : $A \> H \rightarrow H$
| $p_i$ : $\prod (x : B_i \> H), t_i = r_i$ $\quad (i = 1, \ldots, m)$
\end{lstlisting}
where $A$ and each $B_i$ are polynomials, and each $t_i$ and $r_i$ are constructor terms over $c$ of type $H$ with $x : B_i \> H$ as variable.
\end{definition}

Before we can give the rules for higher inductive types, we need to define the lift of a constructor term.
\begin{definition}
Given is a constructor $c : A \> H \rightarrow H$, a type family $Y : H \rightarrow \Type$, and a term $f : \prod (x: A \> H), \bar{A} \> Y \> x \rightarrow  Y(c \> x)$.
For a constructor term $F \> H \Vdash r : G \> H$ we define the \emph{lift $\dak{r}$ of $r$} with type 
$
x : F \> H, h_x : \bar{F} \> Y \> x \vdash \dak{r} : \bar{G} \> Y \> r
$
by induction in $r$ as follows.
        \begin{align*}
                \dak{t} & := t
                & \dak{x} & := h_x &
                \dak{c_i \> r} & := f_i \> r \> \dak{r}\\
                \dak{\pi_j \> r} & := \pi_j \> \dak{r} &
                \dak{(r_1, r_2)} & := (\dak{r_1}, \dak{r_2})
                & \dak{\inn_j \> r} & := \dak{r}
        \end{align*}
\end{definition}

With all these notions we can give the introduction, elimination and computation rules of higher inductive types.
The introduction rules for $H$ as given in Definition \ref{def:hit} are
\[
c : A \> H \rightarrow H,
\]
\[
p_i : \prod (x : B_i \> H), t_i = r_i.
\]
We also have an elimination rule for which we use the lifting of constructor terms.
\begin{equation*}
        \AxiomC{$\vdash Y : H \rightarrow \Type$}
        \noLine
        \UnaryInfC{$\vdash f : \prod (x : A \> H), \bar{A} \> Y \> x \rightarrow Y \> (c \> x)$}
%       \noLine
        \AxiomC{$\vdash q_i : \prod (x : B_i \> H) (h_x : \bar{B}_i \> Y \> x), \hatt_i =_{p_j \, x} \hatr_j$}
        \BinaryInfC{$\vdash H\frec(f, q_1, \ldots, q_n) : \prod (x : H), Y \> x$}
        \DisplayProof
\end{equation*}
Let us abbreviate $H\frec(f, q_1, \ldots, q_n)$ by $H\frec$.
The type $H$ also has computation rules for each point $t : A \> H$
\begin{equation*}
H\frec \> (c \> t) \equiv f \> t \> (\bar{A} \> H\frec \> t),
\end{equation*}
and for each $a : B_i \> H$
\begin{equation*}
\apd \> H\frec \> (p_i \> a) \equiv q_i \> a \> (\bar{B}_i \> H\frec \> a).
\end{equation*}
Note that these equalities are definitional rather than propositional.

Let us now give some examples of higher inductive types which will be crucial in this paper.
The first one identifies points in some type, and we call it the \emph{coequalizer}.
\lstset{language=Coq}
\begin{lstlisting}
Inductive $\coeq$ $(A, B : \Type)$ $(f, g : A \rightarrow B)$ :=
| $\inC$ : $B \rightarrow \coeq \> A \> B \> f \> g$
| $\glueC$ : $\prod (a : A), \inC \> (f \> a) = \inC \> (g \> a)$
\end{lstlisting}
The elimination rule for this type is as follows
\begin{equation*}
        \AxiomC{$\vdash Y : \coeq \> A \> B \> f \> g \rightarrow \Type$}
        \noLine
        \UnaryInfC{$\vdash i_Y : \prod (b : B), Y \> (\inC \> b)$}
        %       \noLine
        \AxiomC{$\vdash g_Y : \prod (a : A), \glueC_*(i_Y \> (f \> a)) = i_Y \> (g \> a)$}
        \BinaryInfC{$\vdash \coeq\frec(i_Y, g_Y) : \prod (x : \coeq \> A \> B \> f \> g), Y \> x$}
        \DisplayProof
\end{equation*}
Note the similarities with the definition of the coequalizer in category theory \cite{mac2013categories}.

Next we define a type which identifies two paths in some type which we shall call the \emph{path coequalizer}.
\lstset{language=Coq}
\begin{lstlisting}
Inductive $\pcoeq$ $(A, B : \Type)$ $(p : A \rightarrow \sum (b_1, b_2 : B), (b_1 = b_2) \times (b_1 = b_2))$ :=
| $\inP$ : $B \rightarrow \pcoeq \> A \> B \> f \> g$
| $\glueP$ : $\prod (a : A), \ap \> \inC \> (\pi_1(\pi_3(p \> a))) = \ap \> \inC \> (\pi_2(\pi_3(p \> a)))$
\end{lstlisting}
We will abbreviate $p_i \> a = \pi_i(\pi_3(p \> a))$ for $i = 1, 2$.
Note that this type has 2-paths, so it does not follow the given syntax.
Furthermore, the identified paths used in the parameter are obliged to have the same endpoints.
The introduction and computation rules are as expected, but we will give the elimination rule.
\begin{equation*}
        \AxiomC{$\vdash Y : \pcoeq \> A \> B \> f \> g \rightarrow \Type$}
        \noLine
        \UnaryInfC{$\vdash i_Y : \prod (b : B), Y \> (\inP \> b)$}
        %       \noLine
        \AxiomC{$\vdash g_Y : \prod (a : A), \apd \> i_Y \> (p_1 \> a) = \apd \> i_Y \> (p_2 \> a)$}
        \BinaryInfC{$\vdash \pcoeq\frec(i_Y, g_Y) : \prod (x : \pcoeq \> A \> B \> f \> g), Y \> x$}
        \DisplayProof
\end{equation*}

The last HIT we need is the homotopy colimit, and for that we again are inspired by the definition from category theory.
\lstset{language=Coq}
\begin{lstlisting}
Inductive $\hocolim$ $(F : \mathbb{N} \rightarrow \Type)$ $(f : \prod (n : \mathbb{N}, F \> n \rightarrow F(n+1)))$ :=
| $\inc$ : $\prod (n : \mathbb{N}), F \> n \rightarrow$ $\hocolim$ $F$ $f$
| $\com$ : $\prod (n : \mathbb{N}) (x : F \> n), \inc \> n \> x = \inc \> (n+1) \> (f \> n \> x)$
\end{lstlisting}
Now the elimination rule is as follows.
\begin{equation*}
        \AxiomC{$\vdash Y : \hocolim \> F \> f \rightarrow \Type$}
        \noLine
        \UnaryInfC{$\vdash i_Y : \prod (n : \mathbb{N}) (x : F \> n), Y \> (\inc \> n \> x)$}
        \noLine
        \UnaryInfC{$\vdash c_Y : \prod (n : \mathbb{N}) (x : F \> n), \com_*(i_Y \> n \> x) = i_Y \> (n+1) \> (f \> n \> x)$}
        \UnaryInfC{$\vdash \hocolim\frec(i_Y, c_Y) : \prod (x : \hocolim \> F \> f), Y \> x$}
        \DisplayProof
\end{equation*}

@@

\begin{definition}
Let $A, B$ be types, let $f, g : A \rightarrow B$ be maps, and let $x : A$ be a point of $A$.
Then for all $e : f = g$ we define a path $e \fpath x : f \> x = g \> x$ using path induction which sends $\refl$ to $\refl$.
\end{definition}

\begin{definition}
Given are types $A, B$ and two maps $f, g : A \rightarrow B$.
Then we define the type $\funext$, of which the inhabitants represent proofs of function extensionality, as follows
\[
\funext = (\prod (x : A), f \> x = g \> x) \rightarrow f = g.
\]
\end{definition}

We will assume that we have an inhabitant $\FE : \funext$ such that for $p : \prod (x : A), f \> x = g \> x$ and $x : A$ we have
\[
(\FE \> p) \fpath x = p \> x.
\]
In homotopy type theory function extensionality can be proved by assuming univalence \cite{UnivalenceFE} or by assuming the existence of an interval object \cite{IntervalFE}.

@@

Before giving the definition, let us acquire inspiration by looking in more detail at how an inductive type is constructed.
Such a type $T$ is defined by a constructor $c : A \> T \rightarrow T$ with $A$ polynomial.
Equivalently, we can define the type $T$ with a signature which is a polynomial functor $A$, and then by Ad\'amek's theorem $T$ is the following colimit.
\[
\xymatrix
{
        \zero \ar[r] & A \> \zero  \ar[r] & A(A \> \zero) \ar[r] & \ldots
}
\]
To understand what this does, let us assume that $A \> X = 1 + X$, so that $T = \mathbb{N}$.
This means we have two inclusions  $\xymatrix{1 \ar[r] & 1+X}$ and  $\xymatrix{X \ar[r] & 1+X}$, and we call them $0_C$ and $S_C$ respectively.
At every step we formally add for each $x : X$ a successor $S_C \> x$, and we add $0_C$.
Repeatedly applying this construction to the empty type $\zero$ gives the natural numbers $\mathbb{N}$.

@@

To understand more precisely what should be done, let us consider an example.
\lstset{language=Coq}
\begin{lstlisting}
Inductive $\mathbb{N}_2$ :=
| $0$ : $\mathbb{N}_2$
| $S$ : $\mathbb{N}_2 \rightarrow \mathbb{N}_2$
| $p$ : $S(S \> 0) = 0$
\end{lstlisting}
We build a sequence $F$ of approximations.
The first approximation just has a constructor $0$, and after that we add a constructor for $S \> 0$ to obtain the second approximation.
Continuing this way, we ultimately arrive at the third approximation, which we call $F' \> 3$, where we found inhabitants $0$, $S \> 0$, and $S(S \> 0)$.
Now we can make the first identification, and to do so, we take the following homotopy pushout
\[
\xymatrix
{
        \one + \one \ar[d]_{0 + S(S \> 0)} \ar[r]^-{\delta} & I^1 \ar[d]\\
        F' \> 3\ar[r] & F \> 3
}
\]
to obtain the actual third approximation $F \> 3$, and we continue our construction with that one.
Note that to glue during the $n$th step, we need refer to elements from the $(n-2)$th step.
So, in order to do the identification of the $n$th step, inhabitants from a previous approximation need to be used.

Due to the usage of constructor terms, one always has to go back a fixed number of steps.
By extending the syntax, it might be needed to go back an arbitrary amount of steps.
This happens in the following example.
\lstset{language=Coq}
\begin{lstlisting}
Inductive $H$ :=
| $c$ : $H$
| $f$ : $H \rightarrow H$
| $p$ : $\prod (n : \mathbb{N}), f^n \> 0 = 0$
\end{lstlisting}
This example might seem like it is permitted by the syntax, but actually it is not.
The function $f^n$ is defined as a polymorphic map, and using it requires the type $H$ as an argument.
In constructor terms one cannot use the type, and hence this definition is not allowed
An extension with such types will make the construction more complicated, and thus they will not be considered in this paper.

Let us make this idea formal, and for that we start with a higher inductive type given as follows.
\lstset{language=Coq}
\begin{lstlisting}[label={lst:HIT}]
Inductive $H$ :=
| $c$ : $A_{\nonrec} \rightarrow H$
| $c_{\rec}$ : $A_{\rec} \> H \rightarrow H$
| $p_i$ : $\prod (x : B_i \> H), t_i = r_i$ $\quad (i = 1, \ldots, m)$
\end{lstlisting}
Note that the nonrecursive and recursive point constructors are separated in this definition.
The first approximation will be given using the nonrecursive constructors.
\lstset{language=Coq}
\begin{lstlisting}
Inductive $H_{\nonrec}$ :=
| $c_{\nonrec}'$ : $A_{\nonrec} \rightarrow H_{\nonrec}$
\end{lstlisting}

Next we need to generate the other approximations, and that will be done in two steps.
First, we note that types can be extended with a recursive constructor.
\lstset{language=Coq}
\begin{lstlisting}
Inductive $H_{\rec}$ $(P : \Type)$:=
| $c_{\rec}'$ : $A_{\rec} \> P \rightarrow H_{\rec} \> P$
\end{lstlisting}
To do the identifications, we need to be able to interpret the constructor terms.
For that we use that each constructor terms only uses each constructor  a finite amount of times, and thus there is a maximum number $n \geq 1$ of times a constructor is used.
In order to define the approximator, we start with a type $H_{\Con}^n$ in which all the constructor terms can be interpreted.
\lstset{language=Coq}
\begin{lstlisting}
Inductive $H_{\Con}^n$ $(P : \Type)$:=
| $\term$ : $P + H_{\rec} \> P + \ldots + H_{\rec}^n \> P \rightarrow H_{\Con}^n \> P$
\end{lstlisting}

Note that this type is actually a labeled sum.
This is because $H_{\rec} \> P$ is isomorphic to $A_{\rec} \> P$, and thus this $H_{\Con}^n \> P$ is actually $P + A_{\rec} \> P + \ldots + A_{\rec}^n \> P$.
Hence, we do not need inductive types in general for the construction, but we just need to have sums and products.
If $n$ is clear from the context, then we shall not write it down.
From Lemma \ref{lem:constrTerm} we can conclude that we can interpret the constructor terms in $H_{\Con} \> P$

>>>>

\newcommand{\fpath}[0]{: \Box :}
\newcommand{\conc}[0]{\bullet}
\newcommand{\comp}[0]{\circ \,}

@@

For example, if $\nat$ is the type of natural numbers, the expression
$(X + \nat) \times X$ is a polynomial expression in~$X$. We write $P[X]$ to indicate a
polynomial in variable $X$. Given a type $A$, we may substutute $A$ for $X$ to obtain a
type $P[A]$, as follows:
%
\begin{align*}
  X[A] &\defeq A, &
  (P_1 \times P_2)[A] &\defeq P_1[A] \times P_2[A], \\
  C[A] &\defeq C, &
  (P_1 + P_2)[A] &\defeq P_1[A] + P_2[A].
\end{align*}
%
In a similar fashion a map $f : A \to B$ is transformed to a map $P[f] : P[A] \to P[B]$:
%
\begin{align*}
  X[f] &\defeq f, &
  (P_1 \times P_2)[f] \, (x,y) &\defeq (P_1[f]\, x, P_2[f] \, y), \\
  C[f] &\defeq \idmap{C}, &
  (P_1 + P_2)[f](\inl{x}) &\defeq \inl{(P_1[f]\,x)}, \\
  & &
  (P_1 + P_2)[f](\inr{y}) &\defeq \inr{(P_2[f]\,y)}.
\end{align*}
%
Somewhat less obviously, a polynomial acts on a type family $B : A \to \Type$ to
give its lifting $\lift{P}[B] : P[A] \to \Type$,
%
\begin{align*}
  \lift{X}[B] \, u &\defeq B\,u, &
  \lift{P_1 \times P_2}[B] \, u \defeq (\lift{P_1}[B]\,u) \times (\lift{P_2}[B]\,u), \\
  \lift{C}[B] \, u &\defeq C, &
  \lift{P_1 + P_2}[B] \, (\inl{u})\defeq \lift{P_1}[B]\,u, \\
  & &
  \lift{P_1 + P_2}[B] \, (\inr{u})\defeq \lift{P_2}[B]\,u,
\end{align*}
%
while a map $f : \dprod{x : A} B_1\,x \to B_2\,x$ lifts to
$\lift{P}[f] : \dprod{u : P[A]} \lift{P}[B_1]\,u \to \lift{P}[B_2]\,u$:
%
\begin{align*}
  \lift{X}[f] \, u &\defeq f \, u &
  \lift{P_1 \times P_2}[f] \, (u_1, u_2) &\defeq (\lift{P_1}[f]\,u_1, \lift{P_2}[f]\,u_2) \\
  \lift{C}[f] \, u &\defeq u &
  \lift{P_1 + P_2}[f] \, (\inl{u})\defeq \lift{P_1}[f]\,u, \\
  & &
  \lift{P_1 + P_2}[f] \, (\inr{u})\defeq \lift{P_2}[f]\,u.
\end{align*}
%
There is of course a strong analogy between the action of $P[X]$ and a polynomial functor,
but we hesitate to call $P[X]$ a functor as there is no category to speak of, and in any
case the manipulations we have just described are syntactic.

Next we define the notion of a constructor term.
\begin{definition}
\label{def:constructor-term}
Given is a function $c : A : T \rightarrow T$ with $A$ polynomial and $T$ a type.
Then we say $t$ is a \emph{constructor term} over $c$ if we can find polynomials $F$ and $G$ such that $x : F : T \Vdash t : G : T$ can be derived using the following rules.
\begin{gather*}
        \AxiomC{$t : B$}
        \AxiomC{$T$ does not occur in $B$}
        \BinaryInfC{$x:F : T \Vdash t : B$}
        \bottomAlignProof
        \DisplayProof 
         \quad
        \AxiomC{}
        \UnaryInfC{$x : F : T \Vdash x : F : T$}
        \bottomAlignProof
        \DisplayProof
        \quad
        \AxiomC{$x : F : T \Vdash r : A : T$}
        \UnaryInfC{$x : F : T \Vdash c : r : T$}
        \bottomAlignProof
        \DisplayProof
        \\[7pt]
        \AxiomC{$j \in \{1, 2\}$}
        \AxiomC{$x : F : T \Vdash r : G_1 : T \times G_2 : T$}
        \BinaryInfC{$x : F : T \Vdash\pi_j : r : G_j : T$}
        \bottomAlignProof
        \DisplayProof
        \quad
        \AxiomC{$j \in \{1, 2\}$}
        \AxiomC{$x : F : T \Vdash r_j : G_j$}
        \BinaryInfC{$x : F : T \Vdash (r_1, r_2) : G_1 : T\times G_2 : T$}
        \bottomAlignProof
        \DisplayProof
        \\[7pt]
        \AxiomC{$j \in \{1, 2\}$}
        \AxiomC{$x : F : T \Vdash r : G_j : T$}
        \BinaryInfC{$x : F : T : \inn_j : r : G_1 : T + G_2 : T$}
        \bottomAlignProof
        \DisplayProof
\end{gather*}
\end{definition}

Using constructor terms we give the following scheme of higher inductive types.
\begin{definition}
\label{def:hit}
A \emph{higher inductive type} is defined according to the following scheme
\lstset{language=Coq}
\begin{lstlisting}
Inductive $H$ :=
| $c$ : $A : H \rightarrow H$
| $p_i$ : $\prod (x : B_i : H), t_i = r_i$ $\quad (i = 1, \ldots, m)$
\end{lstlisting}
where $A$ and each $B_i$ are polynomials, and each $t_i$ and $r_i$ are constructor terms over $c$ of type $H$ with $x : B_i : H$ as variable.
\end{definition}

Before we can give the rules for higher inductive types, we need to define the lift of a constructor term.
\begin{definition}
Given is a constructor $c : A : H \rightarrow H$, a type family $Y : H \rightarrow \Type$, and a term $f : \prod (x: A : H), \bar{A} : Y : x \rightarrow  Y(c : x)$.
For a constructor term $F : H \Vdash r : G : H$ we define the \emph{lift $\dak{r}$ of $r$} with type 
$
x : F : H, h_x : \bar{F} : Y : x \vdash \dak{r} : \bar{G} : Y : r
$
by induction in $r$ as follows.
        \begin{align*}
                \dak{t} & := t
                & \dak{x} & := h_x &
                \dak{c_i : r} & := f_i : r : \dak{r}\\
                \dak{\pi_j : r} & := \pi_j : \dak{r} &
                \dak{(r_1, r_2)} & := (\dak{r_1}, \dak{r_2})
                & \dak{\inn_j : r} & := \dak{r}
        \end{align*}
\end{definition}

With all these notions we can give the introduction, elimination and computation rules of higher inductive types.
The introduction rules for $H$ as given in Definition \ref{def:hit} are
\[
c : A : H \rightarrow H,
\]
\[
p_i : \prod (x : B_i : H), t_i = r_i.
\]
We also have an elimination rule for which we use the lifting of constructor terms.
\begin{equation*}
        \AxiomC{$\vdash Y : H \rightarrow \Type$}
        \noLine
        \UnaryInfC{$\vdash f : \prod (x : A : H), \bar{A} : Y : x \rightarrow Y : (c : x)$}
%       \noLine
        \AxiomC{$\vdash q_i : \prod (x : B_i : H) (h_x : \bar{B}_i : Y : x), \hatt_i =_{p_j \, x} \hatr_j$}
        \BinaryInfC{$\vdash H\frec(f, q_1, \ldots, q_n) : \prod (x : H), Y : x$}
        \DisplayProof
\end{equation*}
Let us abbreviate $H\frec(f, q_1, \ldots, q_n)$ by $H\frec$.
The type $H$ also has computation rules for each point $t : A : H$
\begin{equation*}
H\frec : (c : t) \equiv f : t : (\bar{A} : H\frec : t),
\end{equation*}
and for each $a : B_i : H$
\begin{equation*}
\apd : H\frec : (p_i : a) \equiv q_i : a : (\bar{B}_i : H\frec : a).
\end{equation*}
Note that these equalities are definitional rather than propositional.

Let us now give some examples of higher inductive types which will be crucial in this paper.
The first one identifies points in some type, and we call it the \emph{coequalizer}.
\lstset{language=Coq}
\begin{lstlisting}
Inductive $\coeq$ $(A, B : \Type)$ $(f, g : A \rightarrow B)$ :=
| $\inC$ : $B \rightarrow \coeq : A : B : f : g$
| $\glueC$ : $\prod (a : A), \inC : (f : a) = \inC : (g : a)$
\end{lstlisting}
The elimination rule for this type is as follows
\begin{equation*}
        \AxiomC{$\vdash Y : \coeq : A : B : f : g \rightarrow \Type$}
        \noLine
        \UnaryInfC{$\vdash i_Y : \prod (b : B), Y : (\inC : b)$}
        %       \noLine
        \AxiomC{$\vdash g_Y : \prod (a : A), \glueC_*(i_Y : (f : a)) = i_Y : (g : a)$}
        \BinaryInfC{$\vdash \coeq\frec(i_Y, g_Y) : \prod (x : \coeq : A : B : f : g), Y : x$}
        \DisplayProof
\end{equation*}
Note the similarities with the definition of the coequalizer in category theory \cite{mac2013categories}.

Next we define a type which identifies two paths in some type which we shall call the \emph{path coequalizer}.
\lstset{language=Coq}
\begin{lstlisting}
Inductive $\pcoeq$ $(A, B : \Type)$ $(p : A \rightarrow \sum (b_1, b_2 : B), (b_1 = b_2) \times (b_1 = b_2))$ :=
| $\inP$ : $B \rightarrow \pcoeq : A : B : f : g$
| $\glueP$ : $\prod (a : A), \ap : \inC : (\pi_1(\pi_3(p : a))) = \ap : \inC : (\pi_2(\pi_3(p : a)))$
\end{lstlisting}
We will abbreviate $p_i : a = \pi_i(\pi_3(p : a))$ for $i = 1, 2$.
Note that this type has 2-paths, so it does not follow the given syntax.
Furthermore, the identified paths used in the parameter are obliged to have the same endpoints.
The introduction and computation rules are as expected, but we will give the elimination rule.
\begin{equation*}
        \AxiomC{$\vdash Y : \pcoeq : A : B : f : g \rightarrow \Type$}
        \noLine
        \UnaryInfC{$\vdash i_Y : \prod (b : B), Y : (\inP : b)$}
        %       \noLine
        \AxiomC{$\vdash g_Y : \prod (a : A), \apd : i_Y : (p_1 : a) = \apd : i_Y : (p_2 : a)$}
        \BinaryInfC{$\vdash \pcoeq\frec(i_Y, g_Y) : \prod (x : \pcoeq : A : B : f : g), Y : x$}
        \DisplayProof
\end{equation*}

The last HIT we need is the homotopy colimit, and for that we again are inspired by the definition from category theory.
\lstset{language=Coq}
\begin{lstlisting}
Inductive $\hocolim$ $(F : \mathbb{N} \rightarrow \Type)$ $(f : \prod (n : \mathbb{N}, F : n \rightarrow F(n+1)))$ :=
| $\inc$ : $\prod (n : \mathbb{N}), F : n \rightarrow$ $\hocolim$ $F$ $f$
| $\com$ : $\prod (n : \mathbb{N}) (x : F : n), \inc : n : x = \inc : (n+1) : (f : n : x)$
\end{lstlisting}
Now the elimination rule is as follows.
\begin{equation*}
        \AxiomC{$\vdash Y : \hocolim : F : f \rightarrow \Type$}
        \noLine
        \UnaryInfC{$\vdash i_Y : \prod (n : \mathbb{N}) (x : F : n), Y : (\inc : n : x)$}
        \noLine
        \UnaryInfC{$\vdash c_Y : \prod (n : \mathbb{N}) (x : F : n), \com_*(i_Y : n : x) = i_Y : (n+1) : (f : n : x)$}
        \UnaryInfC{$\vdash \hocolim\frec(i_Y, c_Y) : \prod (x : \hocolim : F : f), Y : x$}
        \DisplayProof
\end{equation*}

@@

\begin{definition}
Let $A, B$ be types, let $f, g : A \rightarrow B$ be maps, and let $x : A$ be a point of $A$.
Then for all $e : f = g$ we define a path $e \fpath x : f : x = g : x$ using path induction which sends $\refl$ to $\refl$.
\end{definition}

\begin{definition}
Given are types $A, B$ and two maps $f, g : A \rightarrow B$.
Then we define the type $\funext$, of which the inhabitants represent proofs of function extensionality, as follows
\[
\funext = (\prod (x : A), f : x = g : x) \rightarrow f = g.
\]
\end{definition}

We will assume that we have an inhabitant $\FE : \funext$ such that for $p : \prod (x : A), f : x = g : x$ and $x : A$ we have
\[
(\FE : p) \fpath x = p : x.
\]
In homotopy type theory function extensionality can be proved by assuming univalence \cite{UnivalenceFE} or by assuming the existence of an interval object \cite{IntervalFE}.

@@

Before giving the definition, let us acquire inspiration by looking in more detail at how an inductive type is constructed.
Such a type $T$ is defined by a constructor $c : A : T \rightarrow T$ with $A$ polynomial.
Equivalently, we can define the type $T$ with a signature which is a polynomial functor $A$, and then by Ad\'amek's theorem $T$ is the following colimit.
\[
\xymatrix
{
        \zero \ar[r] & A : \zero  \ar[r] & A(A : \zero) \ar[r] & \ldots
}
\]
To understand what this does, let us assume that $A : X = 1 + X$, so that $T = \mathbb{N}$.
This means we have two inclusions  $\xymatrix{1 \ar[r] & 1+X}$ and  $\xymatrix{X \ar[r] & 1+X}$, and we call them $0_C$ and $S_C$ respectively.
At every step we formally add for each $x : X$ a successor $S_C : x$, and we add $0_C$.
Repeatedly applying this construction to the empty type $\zero$ gives the natural numbers $\mathbb{N}$.

@@

To understand more precisely what should be done, let us consider an example.
\lstset{language=Coq}
\begin{lstlisting}
Inductive $\mathbb{N}_2$ :=
| $0$ : $\mathbb{N}_2$
| $S$ : $\mathbb{N}_2 \rightarrow \mathbb{N}_2$
| $p$ : $S(S : 0) = 0$
\end{lstlisting}
We build a sequence $F$ of approximations.
The first approximation just has a constructor $0$, and after that we add a constructor for $S : 0$ to obtain the second approximation.
Continuing this way, we ultimately arrive at the third approximation, which we call $F' : 3$, where we found inhabitants $0$, $S : 0$, and $S(S : 0)$.
Now we can make the first identification, and to do so, we take the following homotopy pushout
\[
\xymatrix
{
        \one + \one \ar[d]_{0 + S(S : 0)} \ar[r]^-{\delta} & I^1 \ar[d]\\
        F' : 3\ar[r] & F : 3
}
\]
to obtain the actual third approximation $F : 3$, and we continue our construction with that one.
Note that to glue during the $n$th step, we need refer to elements from the $(n-2)$th step.
So, in order to do the identification of the $n$th step, inhabitants from a previous approximation need to be used.

Due to the usage of constructor terms, one always has to go back a fixed number of steps.
By extending the syntax, it might be needed to go back an arbitrary amount of steps.
This happens in the following example.
\lstset{language=Coq}
\begin{lstlisting}
Inductive $H$ :=
| $c$ : $H$
| $f$ : $H \rightarrow H$
| $p$ : $\prod (n : \mathbb{N}), f^n : 0 = 0$
\end{lstlisting}
This example might seem like it is permitted by the syntax, but actually it is not.
The function $f^n$ is defined as a polymorphic map, and using it requires the type $H$ as an argument.
In constructor terms one cannot use the type, and hence this definition is not allowed
An extension with such types will make the construction more complicated, and thus they will not be considered in this paper.

Let us make this idea formal, and for that we start with a higher inductive type given as follows.
\lstset{language=Coq}
\begin{lstlisting}[label={lst:HIT}]
Inductive $H$ :=
| $c$ : $A_{\nonrec} \rightarrow H$
| $c_{\rec}$ : $A_{\rec} : H \rightarrow H$
| $p_i$ : $\prod (x : B_i : H), t_i = r_i$ $\quad (i = 1, \ldots, m)$
\end{lstlisting}
Note that the nonrecursive and recursive point constructors are separated in this definition.
The first approximation will be given using the nonrecursive constructors.
\lstset{language=Coq}
\begin{lstlisting}
Inductive $H_{\nonrec}$ :=
| $c_{\nonrec}'$ : $A_{\nonrec} \rightarrow H_{\nonrec}$
\end{lstlisting}

Next we need to generate the other approximations, and that will be done in two steps.
First, we note that types can be extended with a recursive constructor.
\lstset{language=Coq}
\begin{lstlisting}
Inductive $H_{\rec}$ $(P : \Type)$:=
| $c_{\rec}'$ : $A_{\rec} : P \rightarrow H_{\rec} : P$
\end{lstlisting}
To do the identifications, we need to be able to interpret the constructor terms.
For that we use that each constructor terms only uses each constructor  a finite amount of times, and thus there is a maximum number $n \geq 1$ of times a constructor is used.
In order to define the approximator, we start with a type $H_{\Con}^n$ in which all the constructor terms can be interpreted.
\lstset{language=Coq}
\begin{lstlisting}
Inductive $H_{\Con}^n$ $(P : \Type)$:=
| $\term$ : $P + H_{\rec} : P + \ldots + H_{\rec}^n : P \rightarrow H_{\Con}^n : P$
\end{lstlisting}

Note that this type is actually a labeled sum.
This is because $H_{\rec} : P$ is isomorphic to $A_{\rec} : P$, and thus this $H_{\Con}^n : P$ is actually $P + A_{\rec} : P + \ldots + A_{\rec}^n : P$.
Hence, we do not need inductive types in general for the construction, but we just need to have sums and products.
If $n$ is clear from the context, then we shall not write it down.
From Lemma \ref{lem:constrTerm} we can conclude that we can interpret the constructor terms in $H_{\Con} : P$
